### 2025-07-18 paper/sections/method.qmd
Editorial pass up until S4.2

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address various other desiderata including those introduced above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on various approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in CT.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples. We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged. As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

>>>>

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address many desiderata including the ones discussed above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on the approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in counterfactual training.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. CT adopts gradient-based CE search in training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for the training samples. We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged. As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.
